---
title: "Introduction to ssrfr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to ssrfr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ssrfr)
```

`ssrfr` is under development. It guards one hop at a time:
`ssrf_prepare_hop()` decides whether a URL may be fetched and `ssrf_fetch()`
fetches it through a pinned connection; a redirect is followed by preparing
the next hop from the previous one, or by `ssrf_fetch_chain()`, which runs
that loop. The contract it implements is
`design/specs/ssrfr-v1.md` in the source repository.

Every hop is evaluated under a policy. It is validated when it is built, so a
malformed rule is an error rather than a rule that silently matches nothing:

```{r}
policy <- ssrf_policy(allow_ranges = "10.0.0.0/8", deny_hosts = ".corp")
policy

try(ssrf_policy(deny_hosts = "com, ru"))
```

The reason codes a refusal will carry are a closed, versioned vocabulary:

```{r}
ssrf_vocabulary("reason_codes")$code
```

## Inspecting a URL

`ssrf_inspect_url()` reports what a URL is under a policy, with no network
I/O. It parses the URL as the guarded fetch will, with `rurl` and then with
libcurl's own parser, and reports the first check that applies: a parse the two
parsers disagree on, the scheme, embedded credentials, the port, an ambiguous
numeric host, the hostname rules, and, for a host written as an address, the
address rules.

```{r}
ssrf_inspect_url("http://0177.0.0.1/")$code
ssrf_inspect_url("http://[64:ff9b::a9fe:a9fe]/latest/meta-data/")
```

It is a pre-filter and a way to lint a configuration, not a defense. It
resolves no name, so a name it finds nothing wrong with may still point
anywhere:

```{r}
ssrf_inspect_url("https://example.com/")$code

policy <- ssrf_policy(allow_ranges = "10.0.0.0/8")
ssrf_inspect_url("http://10.1.2.3/", policy)$code
```

The provider endpoints and metadata hostnames it refuses by default are
vocabularies too, each row citing the vendor documentation that names it:

```{r}
ssrf_vocabulary("metadata_hostnames")$hostname
```

## Inspecting what a name resolves to

With `layer = "L1"`, `ssrf_inspect_url()` also asks the system resolver, once,
for every address of a name the earlier checks passed, and classifies each
address with the addresses embedded in it. The name is looked up as absolute,
with a trailing root dot, so `http://intranet/` is never completed with a DNS
search domain. One refused address refuses the whole set, and each address
keeps its own facts. Were `dual.example` to resolve to `93.184.216.34` and
`::1`:

```{r, eval = FALSE}
res <- ssrf_inspect_url("http://dual.example/", layer = "L1")
res$code
#> [1] "loopback"
vapply(res$addresses, function(a) a$code, "")
#> [1] NA         "loopback"
```

A resolver error, an empty answer or an answer that is not an address is the
operational cause `"unresolvable"`, in `cause` rather than `code`. A URL an
earlier check refuses is never looked up, so this makes no DNS query:

```{r}
ssrf_inspect_url("http://files.example:6379/", layer = "L1")$code
```

L1 is still not a defense: the answer can change the moment it returns. Only
the guarded fetch, which connects to an address it has just validated, is.

## Fetching through the guard

A guarded hop has two steps. `ssrf_prepare_hop()` parses the URL, checks it,
resolves a name once and classifies every address, and returns one of three
values: a refusal (`code`), an operational failure (`cause`), or a binding.
Only a binding can be fetched. Nothing below makes a network call: these URLs
are refused, or their host is an address, which needs no resolution.

```{r}
policy <- ssrf_policy()

refusal <- ssrf_prepare_hop("http://169.254.169.254/latest/", policy,
                            request = list())
refusal$code
ssrf_public_reason(refusal)

binding <- ssrf_prepare_hop(
  "https://93.184.216.34/data?page=2#top",
  policy,
  request = list(headers = c(Authorization = "Bearer not-shown"))
)
binding
```

The request plan is part of the decision, not of the fetch: a list of
`method`, `headers`, `body` and `carry`, where `list()` is a plain `GET`. A
header the transport owns, such as `Host`, or a metadata-service marker such
as `Metadata-Flavor`, is an error before anything is parsed:

```{r}
try(ssrf_prepare_hop("https://example.com/", policy,
                     request = list(headers = c(`Metadata-Flavor` = "Google"))))
```

`ssrf_fetch()` then connects only to an address the binding validated. The
connection is pinned with libcurl's `connect_to`, so the `Host` header, SNI
and certificate verification stay bound to the hostname, while the name is
never resolved again. The binding is single-use and spent on entry; the
response owns no connection and prints no body:

```{r, eval = FALSE}
response <- ssrf_fetch(binding)
response
#> <ssrfr_response>
#>   status: 200
#>   type: text/html
#>   body: 1256 bytes (read it with $body)
rawToChar(response$body)
binding$state$status
#> [1] 200
try(ssrf_fetch(binding))
#> Error : This binding was already fetched; prepare the hop again.
```

A failure on the wire is a value, not an error: `cause` is one of
`ssrf_vocabulary("causes")`, and `detail` lists the addresses tried and how
each attempt ended. Like a refusal, project it with `ssrf_public_reason()`
before an untrusted party sees it.

## Following redirects

`ssrf_fetch_chain()` follows a whole redirect chain through the guard and
returns its last outcome: the final response, or the refusal or failure that
ended the chain.

```{r, eval = FALSE}
policy <- ssrf_policy(max_redirects = 5)
result <- ssrf_fetch_chain(
  "https://example.com/start",
  policy,
  request = list(headers = c(Authorization = "Bearer not-shown"))
)
```

It is built on the two primitives and adds nothing to them. A redirect is
returned, not followed, by `ssrf_fetch()`: its status is in the response and
its `Location` in `binding$state$location`. To follow it, prepare the next
hop with `from = binding`. The `Location` may be relative; it is resolved
against the previous hop's URL, and the new URL is checked from the start.
The request plan is inherited, never restated: a redirect to another origin
drops the body and every header field the plan did not nominate in `carry`,
and `Authorization`, `Proxy-Authorization` and `Cookie` never cross. This is
the loop `ssrf_fetch_chain()` runs. It keeps only the last outcome, so an
application that must log every hop, as OWASP's "log all accepted and
blocked network flows" asks, runs the loop itself and logs each hop's
binding and outcome where the comment marks:

```{r, eval = FALSE}
policy <- ssrf_policy(max_redirects = 5)
result <- ssrf_prepare_hop(
  "https://example.com/start",
  policy,
  request = list(headers = c(Authorization = "Bearer not-shown"))
)
while (inherits(result, "ssrfr_binding")) {
  hop <- result
  result <- ssrf_fetch(hop)
  # Log this hop here: `hop` is its spent binding, `result` its outcome.
  followed <- inherits(result, "ssrfr_response") &&
    result$status %in% c(301, 302, 303, 307, 308) &&
    identical(hop$state$location_count, 1L)
  if (!followed) {
    break
  }
  result <- ssrf_prepare_hop(hop$state$location, policy, from = hop)
}
outcome <- if (inherits(result, "ssrfr_response")) {
  result$body # the body's bytes, a raw vector, whatever they hold
} else if (inherits(result, "ssrfr_refusal")) {
  result$code
} else {
  result$cause
}
```

The loop, like `ssrf_fetch_chain()`, ends with `result` holding the chain's
final outcome: a response, a refusal or a failure, whichever call returned
it. A redirect hop can be refused like any other: `"downgrade"` for `https`
to `http`, or any code a first hop can get. Log a refusal's `code` or a
failure's `cause` for the operator, and show an untrusted party only
`ssrf_public_reason(result)`. The `url` of a redirect hop is the `Location`
exactly as `binding$state$location` holds it; a different URL is an error,
and starts a new chain instead. The chain's `max_redirects` and
`total_timeout` are fixed by its first hop: once `max_redirects` redirects
have been followed, `ssrf_fetch()` refuses the next `3xx` as
`"redirect-limit"`, and a policy that restates either budget differently
mid-chain is an error.

## What the guard cannot protect

`ssrfr` guards only requests made through a binding. These stay unguarded,
and an application that lets an attacker influence the URL must not use them
for it:

- base R's `download.file()`, `url()`, and `readLines()`, `read.csv()` or
  `readRDS()` given a URL;
- direct `curl` calls (`curl::curl_fetch_memory()`, `curl::curl()` and the
  rest), and the `curl` command-line tool, whether through `system()` or
  `download.file(method = "curl")`, which also reads `~/.curlrc`;
- the packages that read URLs through them, such as
  `jsonlite::fromJSON(url)`, `data.table::fread(url)`, `xml2::read_xml(url)`
  and `xml2::read_html(url)`.

`ssrfr` is not egress control either: a firewall or a separate network is
the stronger boundary, and `ssrfr` is defense in depth beside it.

## Testing against a local server

A loopback server is itself a refused destination, so a test that fetches
one through the guard reopens loopback in its own policy. There is no test
mode and no switch that turns the guard off; keep this policy in tests:

```{r, eval = FALSE}
test_policy <- ssrf_policy(allow_ranges = "127.0.0.0/8", allow_ports = 8080)
binding <- ssrf_prepare_hop("http://127.0.0.1:8080/health", test_policy,
                            request = list())
ssrf_fetch(binding)$status
```
